Ride at Your Own Risk: Insights into Toronto’s Bicycle Theft Hotspots

### 2024-11-13 paper/paper.Rmd
citations formatted, bike type and value charts added

  <p>In this analysis, we dig into over a decade’s worth of data from the <a href="https://data.torontopolice.on.ca/datasets/TorontoPS::bicycle-thefts-open-data/about" target="_blank" style="color: #FFC107; text-decoration: underline;">Toronto Police Service</a> to answer key questions: When are bikes at the greatest risk? What makes a bike more likely to be targeted? And where are the theft hotspots cyclists should avoid? Each section of our article addresses a different angle of bike theft in Toronto, from seasonal and weekly theft patterns to variations by bicycle type and value, to pinpointing the city’s most high-risk locations.</p>

@@

required_packages <- c("sf", "here", "dplyr", "readr", "purrr", "plotly", "ggplot2", "leaflet", "osmdata", 
                       "ggridges", "cowplot", "lubridate", "tidyverse", "crosstalk", "bookdown", "hrbrthemes", "gganimate")

@@

library(sf)
library(here)
library(dplyr)
library(readr)
library(purrr)
library(plotly)
library(ggplot2)
library(leaflet)
library(osmdata)
library(cowplot)
library(ggridges)
library(bookdown)
library(lubridate)
library(tidyverse)
library(crosstalk)
library(gganimate)
library(hrbrthemes)

@@

<p>This insight could be valuable for both cyclists and advocacy groups like <a href="https://www.cycleto.ca/" target="_blank" style="color: #FFC107; text-decoration: underline;">Cycle Toronto</a>. Understanding that Friday is a peak day for thefts, local authorities could consider increasing surveillance or public awareness efforts on this day, especially in known hotspot areas. Additionally, for cyclists, it suggests that extra caution on Fridays could reduce the risk of theft. The data here underscores the importance of timing in bike theft prevention, showing how risks fluctuate within the week, which may reflect broader social and behavioral patterns around work and leisure in Toronto.</p>\

@@

<div class="main-body">
  <p>The visualization of bike theft and recovery patterns in Toronto @monthly_plot ussing data from 2014 to 2024 reveals a distinct seasonality, with thefts peaking during the warmer months from May to August. This trend likely correlates with <a href="https://www.toronto.ca/wp-content/uploads/2021/04/8f76-2019-Cycling-Public-Option-Survey-City-of-Toronto-Cycling.pdf" target="_blank" style="color: #FFC107; text-decoration: underline;">increase</a> cycling activity during the summer, when more people choose biking as a primary mode of transportation or recreation. Events like <a href="https://www.bikemonth.ca/toronto" target="_blank" style="color: #FFC107; text-decoration: underline;">Bike Month</a> in June, which includes initiatives encouraging cycling, leading to higher exposure of bikes, making them easier targets for theft.</p>

<p>In contrast, the left side of the plot illustrates the number of bikes found each month, which remains consistently low across the years. The low recovery rate highlights systemic issues in bike recovery efforts. Despite <a href="https://www.tps.ca/services/bicycle-registration/" target="_blank" style="color: #FFC107; text-decoration: underline;">city-wide bike registry programs</a> and <a href="https://www.tps.ca/crime-prevention/bicycle-safety-and-theft-prevention/" target="_blank" style="color: #FFC107; text-decoration: underline;">police campaigns</a> aimed at educating cyclists about secure parking and locking techniques, the ratio of bikes recovered to those stolen has remained low. In recent years, Toronto has initiated <a href="https://www.toronto.ca/services-payments/streets-parking-transportation/cycling-in-toronto/bicycle-parking/" style="color: #FFC107; text-decoration: underline;">programs</a> to improve bike parking infrastructure and introduced more secure lock stands in high-theft areas, but these measures appear insufficient given the scale of theft illustrated in this data.</p>

@@

```{r}

@@

<div class="main-body">
  <p>The tallest bar, representing "Bikes Stolen," towers above the others with over <strong style="color: #FFC107;">32,000</strong> reported incidents, emphasizing how prevalent bike theft has become across the city of Toronto. <a href="https://www.ctvnews.ca/canada/should-toronto-tear-up-its-bike-lanes-to-improve-traffic-flow-critics-say-it-s-not-so-simple-1.7103768" style="color: #FFC107; text-decoration: underline;">CTV news</a> have highlighted hotspots, including downtown Toronto and areas near major transit hubs, where cyclists frequently leave bikes unattended. In response, The Toronto Police Service (TPS) has introduced <a href="https://www.tps.ca/media-centre/news-releases/fix-title-project-sentry/" style="color: #FFC107; text-decoration: underline;">Project Sentry</a>, an initiative led by officers in 14 Division aimed at recovering stolen bicycles and reuniting them with their rightful owners across the city. This program is part of TPS’s effort to address the high rates of bike theft and improve recovery outcomes for Toronto cyclists. </p>

 <p>In contrast, the "Bikes Found" bar, with only <strong style="color: #FFC107;">422</strong> cases, reveals a deep gap in recovery efforts. The small size of this bar underscores how difficult it is to retrieve stolen bikes and reunite them with their owners, suggesting a systemic issue in tracking and identifying stolen bicycles. Although Toronto has introduced <a href="https://project529.com/garage" style="color: #FFC107; text-decoration: underline;">Project 529</a> to help track ownership, recovery rates remain low, partly due to limited resources and the challenge of identifying recovered bikes that may have been altered or resold. </p>

@@

<div class="main-body">
  <p>In @map_of_stolen_bike, we dive deeper into the city with a ground-level view—a sea of red dots each marks an individual theft incident. The dots are packed tightly in certain neighborhoods, sprawling across downtown Toronto like a dense red blanket. It paints a vivid picture of the density of bike thefts—evidence that the bustling central part of the city faces a higher risk. Street after street is littered with these red markers, almost as if each streetlight bore witness to a theft. There’s a sharp concentration near transit hubs and popular neighborhoods, suggesting a correlation between foot traffic, bike use, and crime opportunity. The City has been making efforts to combat this issue; for example, Cycle Toronto launched the <a href="https://www.cycleto.ca/council-talking-about-ending-bike-theft-week" target="_blank" style="color: #FFC107; text-decoration: underline;">#EndBikeTheft campaign</a>, advocating for secure parking and better registration systems, with city discussions exploring collaboration with platforms like <a href="https://project529.com/garage" style="color: #FFC107; text-decoration: underline;">529 Garage</a> to help reduce and recover stolen bikes (@citeEndBikeTheft). It’s almost as though you can hear the honks and feel the rush of people in these crowded streets.</p>

@@

# Citations

>>>>

  <p>In this analysis, we dig into over a decade’s worth of data from the <a href="https://data.torontopolice.on.ca/datasets/TorontoPS::bicycle-thefts-open-data/about" target="_blank" style="color: #FFC107; text-decoration: underline;">Toronto Police Service</a> (@citePolice) to answer key questions: When are bikes at the greatest risk? What makes a bike more likely to be targeted? And where are the theft hotspots cyclists should avoid? Each section of our article addresses a different angle of bike theft in Toronto, from seasonal and weekly theft patterns to variations by bicycle type and value, to pinpointing the city’s most high-risk locations.</p>

@@

required_packages <- c("sf", "here", "dplyr", "readr", "purrr", "scales", "plotly", "ggplot2", "leaflet", "osmdata", 
                       "ggridges", "cowplot", "lubridate", "tidyverse", "crosstalk", "bookdown", "hrbrthemes", "gganimate")

@@

library(sf)
library(here)
library(dplyr)
library(readr)
library(purrr)
library(scales)
library(plotly)
library(ggplot2)
library(leaflet)
library(osmdata)
library(cowplot)
library(ggridges)
library(bookdown)
library(lubridate)
library(tidyverse)
library(crosstalk)
library(gganimate)
library(hrbrthemes)

@@

<p>This insight could be valuable for both cyclists and advocacy groups like <a href="https://www.cycleto.ca/" target="_blank" style="color: #FFC107; text-decoration: underline;">Cycle Toronto</a> (@citeEndBikeTheft). Understanding that Friday is a peak day for thefts, local authorities could consider increasing surveillance or public awareness efforts on this day, especially in known hotspot areas. Additionally, for cyclists, it suggests that extra caution on Fridays could reduce the risk of theft. The data here underscores the importance of timing in bike theft prevention, showing how risks fluctuate within the week, which may reflect broader social and behavioral patterns around work and leisure in Toronto.</p>\

@@

<div class="main-body">
  <p>The visualization of bike theft and recovery patterns in Toronto @monthly_plot ussing data from 2014 to 2024 reveals a distinct seasonality, with thefts peaking during the warmer months from May to August. This trend likely correlates with <a href="https://www.toronto.ca/wp-content/uploads/2021/04/8f76-2019-Cycling-Public-Option-Survey-City-of-Toronto-Cycling.pdf" target="_blank" style="color: #FFC107; text-decoration: underline;">increase</a> (@citeSurvey) cycling activity during the summer, when more people choose biking as a primary mode of transportation or recreation. Events like <a href="https://www.bikemonth.ca/toronto" target="_blank" style="color: #FFC107; text-decoration: underline;">Bike Month</a> (@citeBikeMonth) in June, which includes initiatives encouraging cycling, leading to higher exposure of bikes, making them easier targets for theft.</p>

<p>In contrast, the left side of the plot illustrates the number of bikes found each month, which remains consistently low across the years. The low recovery rate highlights systemic issues in bike recovery efforts. Despite <a href="https://www.tps.ca/services/bicycle-registration/" target="_blank" style="color: #FFC107; text-decoration: underline;">city-wide bike registry programs</a> (@citeTPSRegistration) and <a href="https://www.tps.ca/crime-prevention/bicycle-safety-and-theft-prevention/" target="_blank" style="color: #FFC107; text-decoration: underline;">police campaigns</a> (@citeTPSBicycleSafety) aimed at educating cyclists about secure parking and locking techniques, the ratio of bikes recovered to those stolen has remained low. In recent years, Toronto has initiated <a href="https://www.toronto.ca/services-payments/streets-parking-transportation/cycling-in-toronto/bicycle-parking/" style="color: #FFC107; text-decoration: underline;">programs</a> (@citeBicycleParking) to improve bike parking infrastructure and introduced more secure lock stands in high-theft areas, but these measures appear insufficient given the scale of theft illustrated in this data.</p>

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE}

# Load the cleaned dataset
bike_type_counts <- read_csv(here("data", "bike_type_counts.csv"))

# Mapping bike type codes to descriptive names
bike_type_names <- c(
  "Others" = "Other Types",
  "MT" = "Mountain",
  "RG" = "Gravel",
  "OT" = "Other",
  "RC" = "Recreational & Commuting",
  "EL" = "Electric"
)

# Calculate percentages for each bike type
bike_type_counts <- bike_type_counts %>%
  mutate(total = sum(bike_num)) %>%
  group_by(bike_type) %>%
  mutate(percentage = (bike_num / total) * 100)

# Summarize types with less than 5% as "Others"
bike_type_counts <- bike_type_counts %>%
  mutate(bike_type = ifelse(percentage < 5, "Others", bike_type)) %>%
  group_by(bike_type) %>%
  summarise(bike_num = sum(bike_num), percentage = sum(percentage)) %>%
  ungroup()

# Reorder bike types by percentage in descending order
bike_type_counts <- bike_type_counts %>%
  mutate(bike_type = fct_reorder(bike_type, -percentage))

# Create a horizontal stacked bar chart
ggplot(bike_type_counts, aes(x = factor(1), y = percentage, fill = bike_type)) +
  geom_bar(stat = "identity", position = "fill") +
  geom_text(aes(label = paste0(round(percentage, 1), "%")), 
            position = position_fill(vjust = 0.5), size = 3) +
  scale_y_continuous(labels = percent_format()) +
  labs(
    title = "Distribution of Bike Types",
    x = "Bike Types",
    y = "Percentage",
    fill = "Bike Types"
  ) +
scale_fill_manual(labels = bike_type_names, values = c("#FF7043", "#FF8F00", "#FFA726", "#FFB300", "#FFC107", "#FFCA28")) +  coord_flip() +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.position = "right")

```{r, echo=FALSE, message=FALSE, warning=FALSE}

# Load data
bike_cost_counts <- read_csv(here("data", "bike_cost_counts.csv"))

# Create a pie chart of bike cost tiers
ggplot(bike_cost_counts, aes(x = "", y = bike_num, fill = cost_tier)) +
  geom_bar(width = 1, stat = "identity") +
  coord_polar("y") +
  labs(
    title = "Proportion of Bike Cost Tiers",
    fill = "Cost Tier"
  ) +
  scale_fill_manual(values = c("#FFCA28", "#FFC107", "#FFB300", "#FF8F00")) +  # Customize colors
  theme_minimal() +
  theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.text = element_blank(),
    panel.grid = element_blank(),
    plot.title = element_text(hjust = 0.5)
  ) +
  geom_text(aes(label = paste0(round(bike_num / sum(bike_num) * 100, 1), "%")),
            position = position_stack(vjust = 0.5))

@@

<div class="main-body">
  <p>The tallest bar, representing "Bikes Stolen," towers above the others with over <strong style="color: #FFC107;">32,000</strong> reported incidents, emphasizing how prevalent bike theft has become across the city of Toronto. <a href="https://www.ctvnews.ca/canada/should-toronto-tear-up-its-bike-lanes-to-improve-traffic-flow-critics-say-it-s-not-so-simple-1.7103768" style="color: #FFC107; text-decoration: underline;">CTV news</a> (@citeCTV) have highlighted hotspots, including downtown Toronto and areas near major transit hubs, where cyclists frequently leave bikes unattended. In response, The Toronto Police Service (TPS) has introduced <a href="https://www.tps.ca/media-centre/news-releases/fix-title-project-sentry/" style="color: #FFC107; text-decoration: underline;">Project Sentry</a> (@citeTPSProjectSentry), an initiative led by officers in 14 Division aimed at recovering stolen bicycles and reuniting them with their rightful owners across the city. This program is part of TPS’s effort to address the high rates of bike theft and improve recovery outcomes for Toronto cyclists. </p>

 <p>In contrast, the "Bikes Found" bar, with only <strong style="color: #FFC107;">422</strong> cases, reveals a deep gap in recovery efforts. The small size of this bar underscores how difficult it is to retrieve stolen bikes and reunite them with their owners, suggesting a systemic issue in tracking and identifying stolen bicycles. Although Toronto has introduced <a href="https://project529.com/garage" style="color: #FFC107; text-decoration: underline;">Project 529</a> (@cite529) to help track ownership, recovery rates remain low, partly due to limited resources and the challenge of identifying recovered bikes that may have been altered or resold. </p>

@@

<div class="main-body">
  <p>In @map_of_stolen_bike, we dive deeper into the city with a ground-level view—a sea of red dots each marks an individual theft incident. The dots are packed tightly in certain neighborhoods, sprawling across downtown Toronto like a dense red blanket. It paints a vivid picture of the density of bike thefts—evidence that the bustling central part of the city faces a higher risk. Street after street is littered with these red markers, almost as if each streetlight bore witness to a theft. There’s a sharp concentration near transit hubs and popular neighborhoods, suggesting a correlation between foot traffic, bike use, and crime opportunity. The City has been making efforts to combat this issue; for example, Cycle Toronto launched the <a href="https://www.cycleto.ca/council-talking-about-ending-bike-theft-week" target="_blank" style="color: #FFC107; text-decoration: underline;">#EndBikeTheft campaign</a> (@citeEndBikeTheft), advocating for secure parking and better registration systems, with city discussions exploring collaboration with platforms like <a href="https://project529.com/garage" style="color: #FFC107; text-decoration: underline;">529 Garage</a> (@cite529) to help reduce and recover stolen bikes. It’s almost as though you can hear the honks and feel the rush of people in these crowded streets.</p>

@@

<div class="section-heading">
  <span class="section-title-line1">Citations</span><br>
</div>
